---
title: How to aggregate ranking data
date: "2025-01-06"
author: Dan Kelley
output: pdf_document
---

**Abstract.** A scheme is discussed for aggregating rankings made by committee
members. It is argued (with an example) that this is preferable to a simpler
scheme of averaging rankings across committee members. As an alternative to
this memo, see a video explaining the method, at
[https://youtu.be/mOxEfYG6wRg](https://youtu.be/mOxEfYG6wRg).


# Rank aggregation method

The method has two requirements.

1. Each committee member ranks each of the files, with no ties and no gaps.
2. Either there is an odd number of committee members, or one of the members
   has been assigned a "double voting" status.

Assuming these requirements are method, then there is a definitive answer to
any proposed two-candidate ordering. And that means that a so-called
bubble-sort algorithm can be applied to aggregate rankings.

To begin with, suppose there are just 2 candidates, named "Smith" and "Jones".
We start by writing down the names in any order. Using the alphabet, for
example, we might write "Jones" at the top of the list and "Smith" below it.
With this list, ask each committee member if they agree with the ranking. Given
conditions 1 and 2, there is certainly a majority opinion. The core of the
proposed aggregation method is to respect this opinion. After this is done,
there is a definitive ranking of these two candidates.

This can be applied to *any* pair of candidates, always with a clear answer to
the question "should these be reordered". If we work down the list from the top, interchanging names as appropriate, we will end up with the weakest candidate at the bottom.

This procedure is then applied a second time. Of course, the bottom element
need not be considered. At the end of this pass, the bottom two on the list are
in the democratically-decided order.

It should be obvious that this procedure will eventually result in a
rank-ordered list.  For $n$ students, $n-1$ passes are required. In total, the
number of comparisons is proportional to $n^2$, which can of course get large
for a large number of files, but the point is that this procedure is not to be
carried out by actually querying the committee members, because they will have
submitted rankings in numerical form.  A computer code can take care of the
mechanics. Such a code is provided in the next section, with an example in the
section following that.

# R-language code for this method

In the following R code, `names` is a vector of applicant names, and `...` is a
place-holder that indicates that `ranker` can accept an arbitrary number of
rankings, one per committee member.  (This will become clear in the test case
of the next section.)

```{r}
ranker <- function(names, ...) {
    rankings <- data.frame(...)
    m <- ncol(rankings)
    v <- data.frame(names, rankings)
    colnames(v) <- c("Name", paste("R", 1:m, sep = ""))
    vorig <- v
    m <- ncol(rankings)
    criterion <- m / 2
    nexch <- 0
    n <- length(names)
    for (era in seq(1, n - 1)) {
        for (i in seq(1, n - 1)) {
            w <- sapply(seq(2, m + 1), \(j) v[i, j] > v[i + 1, j]) |> sum()
            if (w > criterion) {
                tmp <- v[i, ]
                v[i, ] <- v[i + 1, ]
                v[i + 1, ] <- tmp
                nexch <- nexch + 1
            }
        }
    }
    data.frame(vorig, Rank = as.vector(sapply(names, \(name) which(v[, 1] == name))))
}
```

# Test case

In a real-world application, the ranking data will be stored in a file, but for
illustration, let's store some data in a string instead.

```{r}
dataText <- "
Name R1 R2 R3
   A  3  2  4
   B  5  4  5
   C  1  3  1
   D  4  1  3
   E  2  5  2
"
```
Now we may read the data from that string.

```{r}
d <- read.table(text = dataText, header = TRUE)
```

Note that we have 3 committee members, producing these rankings called `R1`,
`R2` and `R3` respectively.  To do the rank aggregation, we simply call the
above-defined function. For comparison, we also add an average-across-rankings
column.


```{r}
R <- ranker(d$Name, d$R1, d$R2, d$R3)
R$AvgR1R2R3 <- (R$R1 + R$R2 + R$R3) / 3
R
```

Notice that the average-across-rankings value has ties, so that value cannot be used
for ranking, without further work (and committee discussion).
